### 2024-09-24 paper/paper.qmd
Edited the "paper" quarto document - wrote an introduction and the data section, also included references.

---
title: "Toronto Paramedic Services"
subtitle: "A deeper look into what is driving high EMS demand in Toronto"
author: 
  - Julia Lee
thanks: "Code and data are available at: https://github.com/JuliaJLee/Toronto_Paramedic_Services.git"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(tinytex)
library(knitr)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

# Introduction

"In case of an emergency, please call 911." From building walls to road signs, this is a message that appears almost everywhere, and it instills the idea that no matter the circumstance, there will always be emergency services available to help. However, research shows that this is not the case. A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 @auditorreport. With drastically increasing wait times for those who required EMS (emergency medical services), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.

In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" @auditorreport of Toronto's paramedic services, further research on how to manage EMS demand is needed. Thus, this paper seeks to investigate the factors that seem to be driving EMS demand within Toronto. Using Paramedic Services Incident Data offered by Toronto Paramedic Services, the aim is to identify which types of emergencies (e.g. medical emergencies or vehicle accidents) strongly drive EMS demand. Knowing this can help inform effective strategies that allocate resources to these drivers for better EMS demand management.

With this, this paper first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time. Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to determine any key drivers of the total EMS demand. As a result, ...

In what follows, the paper will define the data that was used within this analysis, provide a detailed account of the results, and put forth a meaningful reflection of the analysis along with its limitations and next steps.

# Data

Paramedic Services Incident Data (2017-2022) is provided by Toronto Paramedic Services [@torontoparamedicservices]. The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all updated by hand to the city of Toronto's Open Data portal [@opentorontodata].

To simulate, test, download, and clean the Paramedic Services Incident Data (2017-2022), the statistical programming language R was used [@citeR]. Specific libraries that assisted the analysis include `tidyverse` [@tidyverse], `dplyr` [@dplyr], `opendatatoronto` [@opendatatoronto], `tinytex` [@tinytex], `ggplot2` [@ggplot2], and `knitr` [@knitr].

The Paramedic Services Incident Data is refreshed annually and contains information on every single incident paramedics have responded to since January 1st to December 31st in 2017 to 2022. For every day of each year, there are multiple records of incidents that have occurred. For every record or incident, the date and time when a paramedic unit was dispatched and the type of incident is included.

```{r}
#| echo: false
#| message: false
#| label: tbl-rawdata
#| tbl-cap: "Paramedic Services Incident Data for 2022"

data <- read_csv(here::here("data/raw_data/raw_data_2022.csv"))

knitr::kable(head(data[, 3:4]), "simple", position = "left")
```

@tbl-rawdata above previews this information for each recorded incident in 2022. For this analysis, the variables of interest within this analysis are the different incident types that are recorded. This dataset lists five different incident types: medical emergencies, vehicle accidents, emergency transfers, fires, and airport standbys. These types refer to different circumstances under which individuals may call 911 for EMS.

# Results

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References

### 2024-09-24 paper/paper.qmd
Edited table in data section - added results graph and description.

@@
 library(knitr)
-
-data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+library(ggplot2)
 ```
@@
 For every day of each year, there are multiple records of incidents that have occurred.
-For every record or incident, the date and time when a paramedic unit was dispatched and the type of incident is included.
+For every record, there is an ID number, the date and time when a paramedic unit was dispatched, the type of incident, the severity of the incident (priority), the number of paramedic units at the scene, and the general location of the incident.
+@tbl-rawdata below previews this information for recorded incidents in 2022.
 
@@
 
-knitr::kable(head(data[, 3:4]), "simple", position = "left")
+data <- 
+ data |> 
+  select(-...1)
+
+data |>
+  slice(1:6) |>
+  kable(
+    col.names = c("ID", "Dispatch Time", "Incident Type", "Priority", 
+                  "Units Arrived", "Location"),
+    digits = 1,
+    booktabs = TRUE,
+    linesep = "",
+    align = c("l", "l", "l", "l", "l", "l"),
+    format.args = list(big.mark = ",")
+  )
+
 ```
 
-@tbl-rawdata above previews this information for each recorded incident in 2022.
 For this analysis, the variables of interest within this analysis are the different incident types that are recorded.
@@
 # Results
+
+## EMS Demand from 2017 to 2022
+
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-totalemsdemand
+#| fig-cap: Total EMS Demand for Each Year between 2017 and 2022
+
+data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
+  geom_line(size = 0.5, color="black") + geom_point(colour = "black") +
+  ylim(200000,300000) + theme_light() + 
+  ggtitle("EMS Demand From 2017 to 2022)") +
+  xlab("Year") + ylab("Demand for EMS") + 
+  theme(plot.title = element_text(hjust = 0.5)) 
+```
+@fig-totalemsdemand displays the total number of emergency calls for each year.
+The total number of emergency calls appears to increase in 2020 after a notable decrease in 2019.
+This pattern likely reflects the the onset of the covid-19 pandemic.
+Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS within Toronto.
+
+To further unpack what is driving this consistent demand, each individual incident type is also considered.
+
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-incidentemsdemand
+#| fig-cap: EMS Demand for Each Year between 2017 and 2022
+
+data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+ggplot(data, aes(x = Year)) + 
+  geom_line(aes(y = Total_Emergency_Calls), size = 0.5, colour = "black") +
+  geom_line(aes(y = Medical_Emergency_Calls), size = 0.5, colour = "blue") +
+  geom_line(aes(y = Vehicle_Accident_Calls), size = 0.5, colour = "red") +
+  geom_line(aes(y = Emergency_Transfer_Calls), size = 0.5, colour = "green") +
+  geom_line(aes(y = Fire_Calls), size = 0.5, colour = "orange") +
+  geom_line(aes(y = Airport_Standby_Calls), size = 0.5, colour = "purple")
+```
 

### 2024-09-24 paper/paper.qmd
uploading paper edits and prd file again

@@
 ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
-  geom_line(size = 0.5, color="black") + geom_point(colour = "black") +
+  geom_line(linewidth = 0.5, color="black") + geom_point(colour = "black") +
   ylim(200000,300000) + theme_light() + 

### 2024-09-25 paper/paper.qmd
Updated paper.qmd discussion and abstract. Also, added references to the reference list and rendered pdf.

--- 2024-09-24 paper/paper.qmd
@@
 title: "Toronto Paramedic Services"
-subtitle: "A deeper look into what is driving high EMS demand in Toronto"
+subtitle: "A deeper look into what is driving Emergency Medical Services (EMS) demand in Toronto"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "In 2023, there were over 1,000 cases where an ambulance was unavailable to respond to an emergency call.
+This calls for urgent action because in emergency situations, every moment is critical.
+By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand within Toronto to help inform more effective strategies for managing EMS demand.
+In short, medical emergencies and vehicle accidents appear to be the most prominent drivers of EMS demand, suggesting that strategies that seek to better manage demand in these areas may enhance the overall efficiency of paramedic services."
 format: pdf
@@
 library(ggplot2)
+library(dplyr)
 ```
@@
 A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 @auditorreport.
-With drastically increasing wait times for those who required EMS (emergency medical services), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
+With drastically increasing wait times for those who required Eemergency Medical Services (EMS), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
 
@@
 With this, this paper first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time.
-Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to determine any key drivers of the total EMS demand.
-As a result, ...
+Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to determine which of these five incident types are key drivers of the total EMS demand.
+As a result, this analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.
 
-In what follows, the paper will define the data that was used within this analysis, provide a detailed account of the results, and put forth a meaningful reflection of the analysis along with its limitations and next steps.
+In what follows, the paper will define the data that was used within this analysis, provide a detailed account of the results, and put forth a discussion of the analysis along with its limitations and next steps.
 
@@
 ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
-  geom_line(size = 0.5, color="black") + geom_point(colour = "black") +
+  geom_line(linewidth = 0.5, color="black") + geom_point(colour = "black") +
   ylim(200000,300000) + theme_light() + 
-  ggtitle("EMS Demand From 2017 to 2022)") +
-  xlab("Year") + ylab("Demand for EMS") + 
+  ggtitle("EMS Demand From 2017 to 2022") +
+  labs(x = "Year", y = "Demand for EMS", caption = "Data Source: Toronto Paramedic Services") + 
   theme(plot.title = element_text(hjust = 0.5)) 
@@
 
-To further unpack what is driving this consistent demand, each individual incident type is also considered.
+## EMS Demand from 2017 to 2022 Based on Incident Type
+
+To further unpack what is driving this consistent demand, the calls for individual incident type are also considered.
+@fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls.
+Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are the key drivers that propel EMS demand.
+
+By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
+With @fig-incidentemsdemandzoom, it becomes clear that a second prominent driver of EMS demand is vehicle accidents, implying that a fairly large portion of the emergency calls that are dispatched to paramedic services comes from incidents on the road.
 
@@
 
-ggplot(data, aes(x = Year)) + 
-  geom_line(aes(y = Total_Emergency_Calls), size = 0.5, colour = "black") +
-  geom_line(aes(y = Medical_Emergency_Calls), size = 0.5, colour = "blue") +
-  geom_line(aes(y = Vehicle_Accident_Calls), size = 0.5, colour = "red") +
-  geom_line(aes(y = Emergency_Transfer_Calls), size = 0.5, colour = "green") +
-  geom_line(aes(y = Fire_Calls), size = 0.5, colour = "orange") +
-  geom_line(aes(y = Airport_Standby_Calls), size = 0.5, colour = "purple")
+data |>
+  pivot_longer(
+    cols = c("Total_Emergency_Calls", "Medical_Emergency_Calls", "Vehicle_Accident_Calls", "Emergency_Transfer_Calls", "Fire_Calls", "Airport_Standby_Calls"),
+    names_to = "series",
+    values_to = "value"
+  ) |>
+  ggplot(mapping = aes(x = Year, y = value, color = series)) +
+  geom_line() +
+  geom_point() +
+  theme_light() +
+  labs(
+    x = "Year", color = "Incident Type",
+    caption = "Data Source: Toronto Paramedic Services"
+  ) +
+  scale_color_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
+  ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
+  theme(plot.title = element_text(hjust = 0.5)) +
+  ylim(0,300000) + scale_y_continuous(name = "Demand for EMS", labels = scales::comma)
 ```
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-incidentemsdemandzoom
+#| fig-cap: EMS Demand for Each Year between 2017 and 2022
+
+data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+data |>
+  pivot_longer(
+    cols = c("Vehicle_Accident_Calls", "Emergency_Transfer_Calls", "Fire_Calls", "Airport_Standby_Calls"),
+    names_to = "series",
+    values_to = "value"
+  ) |>
+  ggplot(mapping = aes(x = Year, y = value, color = series)) +
+  geom_line() +
+  geom_point() +
+  theme_light() +
+  labs(
+    x = "Year", color = "Incident Type",
+    caption = "Data Source: Toronto Paramedic Services"
+  ) +
+  scale_color_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Vehicle Accidents")) +
+  ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
+  theme(plot.title = element_text(hjust = 0.5)) +
+  ylim(0,20000) + scale_y_continuous(name = "Demand for EMS", labels = scales::comma)
+```
+
+\newpage
 
@@
 
-## First discussion point {#sec-first-point}
+As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it.
+To this, the given analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport].
+With medical emergencies, a shortage of doctors has placed pressure on paramedic services as residents rely on and turn to these services more often [@cbcarticle].
+This in turn, increases in offload delays - the time it takes to offload patients into hospitals.
+Longer offload delays mean longer response times for other incidents.
+Putting all of this together, a strategy to better manage EMS demand for medical emergencies may be to focus on promoting primary-care providers (i.e.
+family doctors) and to allocate resources to provide additional health services (e.g.
+programs to consult with a health-care professional online from home).
 
-## Second discussion point
+Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
+To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.
 
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+Thus, these findings offer a deeper understanding about the underlying factors that can drive EMS demand, and this understanding can further facilitate the exploration of various strategies that allow paramedic services to effectively operate even under high demand.
 

### 2024-09-25 scripts/03-data_cleaning_FOR_MAP.R
Decided to also include a map of the paramedic services incident data so added another R script that cleans the same data in a way that can be mapped. Also saved the cleaned data in the analysis_data folder. All new documents are labelled with "FOR MAP".

@@ -1,0 +1,107 @@
+#### Preamble ####
+# Purpose: Clean the raw Paramedic Services Incident Data from 2017-2022 to map it
+# Author: Julia Lee
+# Date: 24 September 2024
+# Contact: jlee.lee@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: Download and save the actual Paramedic Services Incident Data from 2017 
+# to 2022 from Open Data Toronto
+# Any other information needed? As the downloaded raw data is saved into separate files, 
+# need to clean each file and then merge them together
+
+
+#### Setting Up the Workspace ####
+library(tidyverse)
+library(dplyr)
+
+#### Cleaning the Data ####
+
+# Want to map the number of calls each forward sortation area had in 2022 for each incident type
+# So, need to take Raw 2022 data and organize it to get a list of forward sortation areas with its 
+# corresponding number of emergency calls 
+
+## Reading in Raw 2022 data  
+
+data <- read_csv("data/raw_data/raw_data_2022.csv")
+#view(data)
+
+## Find unique forward sortation areas (fsa)
+ 
+find_fsa <- unique(data$Forward_Sortation_Area)
+find_fsa
+
+## Organize data by fsa
+
+organized_data <- data[order(data$Forward_Sortation_Area, decreasing = TRUE),]
+view(organized_data)
+
+## As there are many rows with "-" as the forward sortation area, want to remove them
+
+fsa_data <- organized_data[-c(288555:288690),]
+#view(fsa_data)
+
+## Now, want to remove "...1", "ID", "Dispatch_Time", "Priority_Number", and "Units_Arrived_At_Scene" columns
+
+fsa_data_2 <- fsa_data[,-c(1:3,5:6)]
+#view(fsa_data_2)
+
+## Need to create a new dataframe that lists each fsa and its number of incidents for each incident type
+
+### First, create a dataframe that has each fsa in a single row and a list of all the incident types in 
+### a column
+
+fsa_data_3 <- fsa_data_2 %>% 
+  group_by(Forward_Sortation_Area) %>% 
+  summarise(Incident_Type = paste(Incident_Type, collapse = ","))
+#view(fsa_data_3)
+
+### Second, create columns that contain the number of incidents for each incident type for each fsa
+
+fsa_data_3$"Emergency Transfers" <- str_count(fsa_data_3$Incident_Type, "Emergency Transfer")
+#view(fsa_data_3)
+
[…]

### 2024-09-25 scripts/03-data_cleaning_FOR_MAP.R
Edited the preamble of the cleaning R script for mapping

@@ -6,8 +6,7 @@
 # License: MIT
 # Pre-requisites: Download and save the actual Paramedic Services Incident Data from 2017 
 # to 2022 from Open Data Toronto
-# Any other information needed? As the downloaded raw data is saved into separate files, 
-# need to clean each file and then merge them together
+# Any other information needed? N/A
 
 
 #### Setting Up the Workspace ####

### 2024-09-25 paper/paper.qmd
the shape file wouldn't open so tried deleting the folder and replacing it with just the shape file.
tried reading in the shapefile again - did not work
successfully loaded shapefile!
working on the code to map

@@
 
+## Mapping EMS Demand
+
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-map
+#| fig-cap: EMS Demand Across Toronto Forward Sortation Areas (Based on data from 2022)
+
+library(maps)
+library(sf)
+
+# Reading in the shapefile
+toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
+
+# Reading in the data to show in map
+data <- read_csv("data/analysis_data/analysis_data_FOR_MAP.csv")
+
+```
+
 \newpage

### 2024-09-25 paper/paper.qmd
tried importing data, but didn't work - need to try again later
figured out the problem of reading in data for map!

--- 2024-09-25 paper/paper.qmd
@@
 
+## Mapping EMS Demand
+
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-map
+#| fig-cap: EMS Demand Across Toronto Forward Sortation Areas (Based on data from 2022)
+
+library(maps)
+library(sf)
+
+data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+
+# Reading in the shapefile
+#toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
+
+```
+
 \newpage

### 2024-09-26 scripts/03-data_cleaning_FOR_MAP.R
Needed to add a column to cleaned data for map and change a column name. Saved the edited file to the same location (in the "analysis_data" folder with "FOR MAP" in the file name).

@@ -32,7 +32,7 @@ find_fsa
 ## Organize data by fsa
 
 organized_data <- data[order(data$Forward_Sortation_Area, decreasing = TRUE),]
-view(organized_data)
+#view(organized_data)
 
 ## As there are many rows with "-" as the forward sortation area, want to remove them
 
@@ -49,8 +49,8 @@ fsa_data_2 <- fsa_data[,-c(1:3,5:6)]
 ### First, create a dataframe that has each fsa in a single row and a list of all the incident types in 
 ### a column
 
-fsa_data_3 <- fsa_data_2 %>% 
-  group_by(Forward_Sortation_Area) %>% 
+fsa_data_3 <- fsa_data_2 |> 
+  group_by(Forward_Sortation_Area) |> 
   summarise(Incident_Type = paste(Incident_Type, collapse = ","))
 #view(fsa_data_3)
 
@@ -71,6 +71,22 @@ fsa_data_3$"Fires" <- str_count(fsa_data_3$Incident_Type, "Fire")
 fsa_data_3$"Airport Standbys" <- str_count(fsa_data_3$Incident_Type, "Airport Standby")
 view(fsa_data_3)
 
+## Add a column that contains the name of the incident type that had the most demand for each fsa
+
+fsa_data_3$"EMS Demand Driver" <- colnames(fsa_data_3)[apply(fsa_data_3,1,which.max)]
+#view(fsa_data_3)
+
+## Change the "Forward_Sortation_Area" column name to "CFSAUID" - this is the name of the column
+## with the forward sortation areas in the shapefile, so having the same column name can make it 
+## easier to join this cleaned data to the shapefile
+
+fsa_data_3 <-
+  fsa_data_3 |>
+  rename(
+    CFSAUID = Forward_Sortation_Area
+  )
+view(fsa_data_3)
+
 ------------------------------------------------------------------------------------------
 
 #### Saving the Data ####
@@ -89,10 +105,10 @@ write_csv(fsa_data_3, file = "data/analysis_data/analysis_data_FOR_MAP.csv")
 # So, will test if the value for Emergency Transfer calls is equal to the count in the "Incident_Type"
 # column
 
-test_cleaned_data <- read_csv("data/analysis_data/analysis_data_FOR_MAP.csv")
-#view(test_cleaned_data)
+test_clean_map_data <- read_csv("data/analysis_data/analysis_data_FOR_MAP.csv")
+#view(test_clean_map_data)
 
[…]

### 2024-09-26 paper/paper.qmd
Got a working map done - just need to change labels on legend.

--- 2024-09-25 paper/paper.qmd
@@
 ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
-  geom_line(linewidth = 0.5, color="black") + geom_point(colour = "black") +
+  geom_line(linewidth = 0.5, colour="black") + geom_point(colour = "black") +
   ylim(200000,300000) + theme_light() + 
@@
   ) |>
-  ggplot(mapping = aes(x = Year, y = value, color = series)) +
+  ggplot(mapping = aes(x = Year, y = value, colour = series)) +
   geom_line() +
@@
   labs(
-    x = "Year", color = "Incident Type",
+    x = "Year", colour = "Incident Type",
     caption = "Data Source: Toronto Paramedic Services"
   ) +
-  scale_color_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
+  scale_colour_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
   ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
@@
   labs(
-    x = "Year", color = "Incident Type",
+    x = "Year", colour = "Incident Type",
     caption = "Data Source: Toronto Paramedic Services"
   ) +
-  scale_color_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Vehicle Accidents")) +
+  scale_colour_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Vehicle Accidents")) +
   ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
@@
   ylim(0,20000) + scale_y_continuous(name = "Demand for EMS", labels = scales::comma)
+```
+
+## Mapping EMS Demand
+
+```{r}
+#| echo: false 
+#| message: false
+#| label: fig-map
+#| fig-cap: EMS Demand Across Toronto Forward Sortation Areas (Based on data from 2022)
+
+library(maps)
+library(sf)
+
+# Reading in the shapefile
+
+toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
+#view(toronto_forward_sortation_areas)
+
+# Reading in the paramedic services data
+
+map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+#view(map_data)
+
+# Test to see if shapefile produces a map
+
+#ggplot(toronto_forward_sortation_areas) + 
+  #geom_sf()
+
+# Merge map_data and toronto_forward_sortation_areas by the column "CFSAUID"
+
+merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
+#view(merged_map_data)
+
+# Map the merged data
+
+ggplot(merged_map_data) + 
+  geom_sf(aes(fill = "EMS Demand Driver")) +
+  labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
+         x = "Longitude", y="Latitude", 
+       caption = "Data Source: Open Canada 2021 Census - Boundary Files & Toronto Paramedic Services",
+       labels = c("Medical Emergencies")) +
+  theme_light() + theme(plot.title = element_text(hjust = 0.5))
+
 ```

### 2024-09-26 scripts/03-data_cleaning_FOR_MAP.R
Changed formatting (i.e. comments and spacing in the R script) + resaved cleaned data for map.

@@ -73,7 +73,7 @@ view(fsa_data_3)
 
 ## Add a column that contains the name of the incident type that had the most demand for each fsa
 
-fsa_data_3$"EMS Demand Driver" <- colnames(fsa_data_3)[apply(fsa_data_3,1,which.max)]
+fsa_data_3$"EMS_Demand_Driver" <- colnames(fsa_data_3)[apply(fsa_data_3,1,which.max)]
 #view(fsa_data_3)
 
 ## Change the "Forward_Sortation_Area" column name to "CFSAUID" - this is the name of the column

### 2024-09-26 paper/paper.qmd
Finalized map and changed colours of graphs.

@@
   ) +
-  scale_colour_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
+  scale_colour_brewer(palette = "Dark2", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
   ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
@@
   ) +
-  scale_colour_brewer(palette = "Set3", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Vehicle Accidents")) +
+  scale_colour_brewer(palette = "Dark2", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Vehicle Accidents")) +
   ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
@@
 ggplot(merged_map_data) + 
-  geom_sf(aes(fill = "EMS Demand Driver")) +
+  geom_sf(aes(fill = EMS_Demand_Driver)) +
   labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
          x = "Longitude", y="Latitude", 
-       caption = "Data Source: Open Canada 2021 Census - Boundary Files & Toronto Paramedic Services",
-       labels = c("Medical Emergencies")) +
-  theme_light() + theme(plot.title = element_text(hjust = 0.5))
+       caption = "Data Source: Open Canada 2021 Census - Boundary Files & Toronto Paramedic Services") +
+  theme_light() + theme(plot.title = element_text(hjust = 0.5)) +
+  scale_fill_discrete(name = "EMS Demand Driver") +
+  scale_fill_manual(values = c("lightblue"))
 

### 2024-09-26 paper/paper.qmd
Rendered a pdf version of the paper with all visualizations done - now need to edit writing.

@@
 
-```{r}
+```{r, results = 'hide'}
+#| message: false
 #| echo: false 
-#| message: false
+#| warning: false
 #| label: fig-map
@@
          x = "Longitude", y="Latitude", 
-       caption = "Data Source: Open Canada 2021 Census - Boundary Files & Toronto Paramedic Services") +
-  theme_light() + theme(plot.title = element_text(hjust = 0.5)) +
+       caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
+  theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
   scale_fill_discrete(name = "EMS Demand Driver") +

### 2024-09-26 paper/paper.qmd
Updated and edited paper, mainly data section.

@@
 title: "Toronto Paramedic Services"
-subtitle: "A deeper look into what is driving Emergency Medical Services (EMS) demand in Toronto"
+subtitle: "A deeper look into what drives Emergency Medical Services (EMS) demand."
 author: 
@@
 abstract: "In 2023, there were over 1,000 cases where an ambulance was unavailable to respond to an emergency call.
-This calls for urgent action because in emergency situations, every moment is critical.
+This calls for urgent action as every moment is critical in emergency situations.
 By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand within Toronto to help inform more effective strategies for managing EMS demand.
@@
 However, research shows that this is not the case.
-A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 @auditorreport.
-With drastically increasing wait times for those who required Eemergency Medical Services (EMS), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
+A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 [@auditorreport].
+With drastically increasing wait times for those who required Emergency Medical Services (EMS), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
 
-In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" @auditorreport of Toronto's paramedic services, further research on how to manage EMS demand is needed.
+In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed.
 Thus, this paper seeks to investigate the factors that seem to be driving EMS demand within Toronto.
@@
 
-In what follows, the paper will define the data that was used within this analysis, provide a detailed account of the results, and put forth a discussion of the analysis along with its limitations and next steps.
+In what follows, the paper will define the data that was used within this analysis (@sec-data), provide a detailed account of the results (@sec-results), and put forth a discussion of the analysis along with its limitations and next steps (@sec-discussion).
 
-# Data
+# Data {#sec-data}
+
+## Toronto Paramedic Services Data
 
@@
 To simulate, test, download, and clean the Paramedic Services Incident Data (2017-2022), the statistical programming language R was used [@citeR].
-Specific libraries that assisted the analysis include `tidyverse` [@tidyverse], `dplyr` [@dplyr], `opendatatoronto` [@opendatatoronto], `tinytex` [@tinytex], `ggplot2` [@ggplot2], and `knitr` [@knitr].
+Specific libraries that assisted the analysis include `tidyverse` [@tidyverse], `dplyr` [@dplyr], `opendatatoronto` [@opendatatoronto], `tinytex` [@tinytex], `ggplot2` [@ggplot2], `knitr` [@knitr], `maps` [@maps], and `sf` [@sf].
 
@@
 data |>
-  slice(1:6) |>
+  slice(1:3) |>
   kable(
@@
 
-For this analysis, the variables of interest within this analysis are the different incident types that are recorded.
-This dataset lists five different incident types: medical emergencies, vehicle accidents, emergency transfers, fires, and airport standbys.
-These types refer to different circumstances under which individuals may call 911 for EMS.
+Looking at @tbl-rawdata, the variable, "ID", is a number that identifies each incident in accordance with a Computer Aided Dispatch (CAD) record [@torontoparamedicservices].
+"Dispatch Time" represents the date and time, measured in minutes and seconds, at which paramedics responded to an emergency call.
+This data further provides information on the type of emergency paramedics responded to with the variable, "Incident Type".
+"Priority" refers to the urgency of all incoming emergency calls.
+The priority level is measured using a "Medical Priority Dispatch System (MPDS)" that organizes incoming information from 911 callers into different categories.
+For instance, in @tbl-rawdata, the priority level is 1 for all three incidents, meaning that the MPDS assessed the information provided by callers and their situations as extremely urgent.
+"Units Arrived" is a count of the number of paramedic units that were present at the scene.
+Lastly, "Location" characterizes the approximate location of each incident and is measured using Forward Sortation Areas (i.e.
+the first three characters of postal codes).
 
-# Results
+## Analysis Data {#sec-analysis}
+
+For this analysis, the variables of interest are the different types of incidents that are recorded.
+The Paramedic Services Incident Data lists five different incident types: (1) medical emergencies, (2) motor vehicle accidents, (3) emergency transfers, (4) fires, and (5) airport standbys.
+Each type refers to a different circumstance under which individuals may call 911 for EMS.
+
+As the objective is to understand the driving factors of EMS demand, the actual data that is used within this analysis reflects the total EMS demand and the demand for each incident type over the course of six years (2017-2022).
+All recorded incidents were aggregated together by incident type to see the number of emergency calls for each type.
+Then, the number of emergency calls for each incident type were summed to produce a total EMS demand count.
+This process was repeated every year from 2017 to 2022.
+@tbl-analysisdata (found below) illustrates the outcome of this process, and summary statistics for this analysis data can be found in the Appendix (@sec-appendix).
+
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-analysisdata
+#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"
+
+data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+data |>
+  slice(1:6) |>
+  kable(
+    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
+                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
+    digits = 1,
+    booktabs = TRUE,
+    linesep = "",
+    align = c("l", "l", "l", "l", "l", "l")
+  )
+```
+
+The column "Unclassified Calls" in @tbl-analysisdata represents the emergency calls that were not assigned an incident type.
+It is important to note that throughout the analysis, these records are omitted.
+
+## Map Data
+
+To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map the driver of EMS demand within each FSA.
+For the maps within this analysis, the analysis data explained in @sec-analysis needed to be modified.
+By grouping the data by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found.
+The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls.
+A small snippet of the modified analysis data is shown below in @tbl-mapdata.
+
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-mapdata
+#| tbl-cap: "Modified Analysis Data for Mapping"
+
+data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+
+data |>
+  slice(1:2) |>
+  kable(
+    col.names = c("CFSAUID", "Incident Type", "Emergency Transfers", "Medical Emergencies",
+                  "Vehicle Accidents", "Fires", "Airport Standbys", "EMS Demand Driver"),
+    digits = 1,
+    booktabs = TRUE,
+    linesep = "",
+    align = c("r", "r", "r", "r", "c", "r"),
+    format.args = list(big.mark = ",")
+  )
+```
+
+To map this data, a shapefile with the boundaries for Toronto's FSAs was required. This data was acquired from the Open Government Data portal by the Government of Canada (CITATION).
+
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 
+\appendix
+
+# Appendix {#sec-appendix}
+
+## Analysis Data Summary Statistics
+
+```{r}
+#| echo: false
+#| message: false
+#| label: tbl-analysisdata
+#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"
+
+data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+summary_data <- summary(data[1,2:8])
+view(summary_data)
+
+data |>
+  slice(1:6) |>
+  kable(
+    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
+                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
+    digits = 1,
+    booktabs = TRUE,
+    linesep = "",
+    align = c("l", "l", "l", "l", "l", "l"),
+    format.args = list(big.mark = ",")
+  )
+
+```
+
+## Map Data
+
+\newpage
+
 # References

### 2024-09-26 paper/paper.qmd
Edited the data section to include explanations of shapefiles, added an appendix, and added explanations of implications of results.

@@
 This process was repeated every year from 2017 to 2022.
-@tbl-analysisdata (found below) illustrates the outcome of this process, and summary statistics for this analysis data can be found in the Appendix (@sec-appendix).
+@tbl-analysis (found below) illustrates the outcome of this process, and summary statistics for this analysis data can be found in the Appendix (@sec-appendix).
 
@@
 #| message: false
-#| label: tbl-analysisdata
+#| label: tbl-analysis
 #| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"
@@
   kable(
-    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
+    col.names = c("Year", "Medical Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
                   "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
@@
 
-The column "Unclassified Calls" in @tbl-analysisdata represents the emergency calls that were not assigned an incident type.
+The column "Unclassified Calls" in @tbl-analysis represents the emergency calls that were not assigned an incident type.
 It is important to note that throughout the analysis, these records are omitted.
@@
 A small snippet of the modified analysis data is shown below in @tbl-mapdata.
+The column that contains each FSA is named "CFSAUID" to ensure that each FSA can be mapped.
 
@@
 
-To map this data, a shapefile with the boundaries for Toronto's FSAs was required. This data was acquired from the Open Government Data portal by the Government of Canada (CITATION).
+To map this data, a shapefile with the boundaries for Toronto's FSAs was required.
+This data was acquired from the Open Government Data portal by the Government of Canada [@2021census].
+Ambulance station location data was also considered in this analysis.
+This dataset provided by Toronto Paramedic Services [@torontoparamedicservices] and shared through the city of Toronto's Open Data portal [@opentorontodata], contains the exact geographic locations of each ambulance station within Toronto.
+Additional information about these datasets, including how they were prepared and used can be found in the Appendix (@sec-appendix).
 
@@
 ```
+
 @fig-totalemsdemand displays the total number of emergency calls for each year.
@@
 @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls.
-Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are the key drivers that propel EMS demand.
-
-By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
-With @fig-incidentemsdemandzoom, it becomes clear that a second prominent driver of EMS demand is vehicle accidents, implying that a fairly large portion of the emergency calls that are dispatched to paramedic services comes from incidents on the road.
+Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.
 
@@
 ```
+
+By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
+With @fig-incidentemsdemandzoom, it becomes clear that a second driver of EMS demand may be vehicle accidents, though not as strong as medical emergencies.
+
 ```{r}
@@
 
+\newpage
+
+As a prominent driver of EMS demand is medical-emergency-related calls, strategies that focus on efficiently managing calls for medical emergencies may help to better navigate EMS demand as a whole.
+Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport].
+With medical emergencies, a shortage of doctors has placed pressure on paramedic services as residents rely on and turn to these services more often [@cbcarticle].
+This in turn, increases in offload delays - the time it takes to offload patients into hospitals.
+Longer offload delays mean longer response times for other incidents.
+Putting all of this together, a strategy to better manage EMS demand for medical emergencies may be to focus on promoting primary-care providers (i.e.
+family doctors) and to allocate resources to provide additional health services (e.g.
+programs to consult with a health-care professional online from home).
+
+Similarly, as vehicle accidents are a second potential driver of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident-related calls.
+Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
+To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.
+
 ## Mapping EMS Demand
+
+By examining EMS demand through different incident types, the investigation above finds that medical emergencies...
 
@@
 toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
-#view(toronto_forward_sortation_areas)
+view(toronto_forward_sortation_areas)
 
@@
 
-As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it.
+## Summary
+
+As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it. Thus, this analysis investigates...
+
 To this, the given analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.
 
-Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport].
-With medical emergencies, a shortage of doctors has placed pressure on paramedic services as residents rely on and turn to these services more often [@cbcarticle].
-This in turn, increases in offload delays - the time it takes to offload patients into hospitals.
-Longer offload delays mean longer response times for other incidents.
-Putting all of this together, a strategy to better manage EMS demand for medical emergencies may be to focus on promoting primary-care providers (i.e.
-family doctors) and to allocate resources to provide additional health services (e.g.
-programs to consult with a health-care professional online from home).
+Thus, these findings offer a deeper understanding about the underlying factors that can drive EMS demand, and this understanding can further facilitate the exploration of various strategies that allow paramedic services to effectively operate even under high demand.
 
-Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
-To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.
+## Limitations
 
-Thus, these findings offer a deeper understanding about the underlying factors that can drive EMS demand, and this understanding can further facilitate the exploration of various strategies that allow paramedic services to effectively operate even under high demand.
+CONTENT
 
@@
 #| label: tbl-analysisdata
-#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"
+#| tbl-cap: "Paramedic Services Incident Data Summary Statistics"
 
@@
 
-summary_data <- summary(data[1,2:8])
-view(summary_data)
+summary_data <- summary(data[,2:8])
+#view(summary_data)
 
-data |>
-  slice(1:6) |>
+summary_data |>
     kable(
-    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
-                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
+    col.names = c("Medical Emergencies", "Vehicle Accidents", "Emergency Transfers",
+                  "Fires", "Airport Standbys", "Unclassified", "Total Calls"),
     digits = 1,
@@
 
+@tbl-analysisdata presents the summary statistics for each recorded incident type for the years 2017 to 2022.
+
 ## Map Data
+
+The shapefile that contains the boundaries for Toronto's FSAs also contained the FSAs for every province and territory of Canada.
+As this analysis only considers the city of Toronto, ArcGIS Pro software [@arcgis] use employed to remove the spatial information of FSA locations outside Toronto.
+Specifically, the records of FSA regions that were not in Toronto were removed from the shapefile's attribute table (i.e.
+data table) using ArcGIS Pro.
+After modifying this data, it was merged with the data presented in @tbl-mapdata and mapped to show the "EMS Demand Driver" for each FSA.
+The final map is shown in @fig-map.
+
+The ambulance station location data was not modified for this analysis. It was mapped separately and added to @fig-map to create @(FIG NAME FOR SECOND MAP).
 

### 2024-09-26 paper/paper.qmd
Needed to change shapefile for map so deleted shapefile folder + edited paper

@@
 This data was acquired from the Open Government Data portal by the Government of Canada [@2021census].
-Ambulance station location data was also considered in this analysis.
-This dataset provided by Toronto Paramedic Services [@torontoparamedicservices] and shared through the city of Toronto's Open Data portal [@opentorontodata], contains the exact geographic locations of each ambulance station within Toronto.
-Additional information about these datasets, including how they were prepared and used can be found in the Appendix (@sec-appendix).
+Additional information about this dataset, including how it was prepared and used can be found in the Appendix (@sec-appendix).
 
@@
 
-By examining EMS demand through different incident types, the investigation above finds that medical emergencies...
+By examining EMS demand through different incident types, the investigation above finds that medical emergencies are responsible for a significantly large portion of 911 calls.
+However, this finding applies to the city of Toronto in general, and emergency calls for medical emergencies may not be the driver of EMS demand within individual regions of Toronto.
+Thus, the following part of this analysis looks to identify the driver for each Forward Sortation Area (FSA) in Toronto.
+If different FSAs have different drivers of EMS demand, then strategies that are tailored to each region's circumstance may be more effective in managing EMS demand across the city.
+
+@fig-map
 
@@
 library(sf)
+library(repr)
 
@@
 
-toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
-view(toronto_forward_sortation_areas)
+#toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
+#view(toronto_forward_sortation_areas)
 
@@
 
-map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+#map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
 #view(map_data)
@@
 
-merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
+#merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
 #view(merged_map_data)
@@
 
-ggplot(merged_map_data) + 
-  geom_sf(aes(fill = EMS_Demand_Driver)) +
-  labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
-         x = "Longitude", y="Latitude", 
-       caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
-  theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
-  scale_fill_discrete(name = "EMS Demand Driver") +
-  scale_fill_manual(values = c("lightblue"))
+#options(repr.plot.width = 20, repr.plot.height = 16)
 
+#ggplot(merged_map_data) + 
+  #geom_sf(aes(fill = EMS_Demand_Driver)) +
+  #labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
+         #x = "Longitude", y="Latitude", 
+       #caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
+  #theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
+  #scale_fill_discrete(name = "EMS Demand Driver") +
+  #scale_fill_manual(values = c("lightblue")) +
+  #geom_sf_label(mapping = aes(label = merged_map_data$CFSAUID), size = 3, check_overlap = TRUE)
 ```
@@
 
-The ambulance station location data was not modified for this analysis. It was mapped separately and added to @fig-map to create @(FIG NAME FOR SECOND MAP).
-
 \newpage

### 2024-09-26 paper/paper.qmd
Continued with interpretation of map - need to finish map section, edit title, intro, abstract, and discussion.

@@
 To map this data, a shapefile with the boundaries for Toronto's FSAs was required.
-This data was acquired from the Open Government Data portal by the Government of Canada [@2021census].
+This data, "2021 Census - Boundary Files", was acquired from the Open Government Data portal by the Government of Canada [@2021census].
 Additional information about this dataset, including how it was prepared and used can be found in the Appendix (@sec-appendix).
@@
 
-@fig-map
+It is important to note that there are only 96 FSAs that can be mapped for the city of Toronto [@2021census].
+For this reason, while the Paramedic Services Incident Data provides records of emergency calls for 184 unique FSAs, only 96 could be mapped in @fig-map below.
+So, although Toronto Paramedic Services have responded to calls throughout the Census Metropolitan Area (CMA) of Toronto and other surrounding regions, calls that occurred within the city of Toronto as outlined by the 2021 Census - Boundary Files data are considered.
 
@@
 
-#toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
+toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas(TOR)")
 #view(toronto_forward_sortation_areas)
@@
 
-#map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
 #view(map_data)
@@
 
-#merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
+merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
 #view(merged_map_data)
@@
 
-#options(repr.plot.width = 20, repr.plot.height = 16)
+options(repr.plot.width = 20, repr.plot.height = 16)
 
-#ggplot(merged_map_data) + 
-  #geom_sf(aes(fill = EMS_Demand_Driver)) +
-  #labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
-         #x = "Longitude", y="Latitude", 
-       #caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
-  #theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
-  #scale_fill_discrete(name = "EMS Demand Driver") +
-  #scale_fill_manual(values = c("lightblue")) +
-  #geom_sf_label(mapping = aes(label = merged_map_data$CFSAUID), size = 3, check_overlap = TRUE)
+ggplot(merged_map_data) + 
+  geom_sf(aes(fill = EMS_Demand_Driver)) +
+  labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
+         x = "Longitude", y="Latitude", 
+       caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
+  theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
+  scale_fill_discrete(name = "EMS Demand Driver") +
+  scale_fill_manual(values = c("lightblue"))
 ```
+
+@fig-map illustrates the drivers of EMS demand for each FSA within the city of Toronto, and the key driver for all 96 FSAs in the city appears to be medical emergencies.
+This shows that for at least these 96 FSAs, strategies that focus on promoting health and well-being may be effective in reducing the pressure paramedic services face with high EMS demand.
+However, there are many more FSAs that paramedic services respond to and [TABLE] shows the remaining FSAs and their respective EMS demand drivers.
+
+From [TABLE], it is evident that not all FSAs have the same driver for EMS demand.
+For instance, the FSA, "", has emergency transfers as its driver while "" has accidents as it driver.
+Thus, the takeaway from this analysis is that paramedic service strategies that are tailored to a specific region and its characteristics and needs, may be better.
+As each region or FSA will vary from others, this variability will make a "one size fits all approach" diffiuclt to employ.
+so, tailored strategies within paramedic services (allocation of resources, staff) as well as within each region (e.g.
+programs) are crucial.
 

### 2024-09-27 scripts/03-data_cleaning_FOR_MAP.R
Needed to remove two rows in clean data for map + resaved clean data file for map

@@ -46,7 +46,7 @@ fsa_data_2 <- fsa_data[,-c(1:3,5:6)]
 
 ## Need to create a new dataframe that lists each fsa and its number of incidents for each incident type
 
-### First, create a dataframe that has each fsa in a single row and a list of all the incident types in 
+### Create a dataframe that has each fsa in a single row and a list of all the incident types in 
 ### a column
 
 fsa_data_3 <- fsa_data_2 |> 
@@ -54,7 +54,7 @@ fsa_data_3 <- fsa_data_2 |>
   summarise(Incident_Type = paste(Incident_Type, collapse = ","))
 #view(fsa_data_3)
 
-### Second, create columns that contain the number of incidents for each incident type for each fsa
+### Create columns that contain the number of incidents for each incident type for each fsa
 
 fsa_data_3$"Emergency Transfers" <- str_count(fsa_data_3$Incident_Type, "Emergency Transfer")
 #view(fsa_data_3)
@@ -69,7 +69,7 @@ fsa_data_3$"Fires" <- str_count(fsa_data_3$Incident_Type, "Fire")
 #view(fsa_data_3)
 
 fsa_data_3$"Airport Standbys" <- str_count(fsa_data_3$Incident_Type, "Airport Standby")
-view(fsa_data_3)
+#view(fsa_data_3)
 
 ## Add a column that contains the name of the incident type that had the most demand for each fsa
 
@@ -85,6 +85,11 @@ fsa_data_3 <-
   rename(
     CFSAUID = Forward_Sortation_Area
   )
+#view(fsa_data_3)
+
+## Remove 2 rows with fsa that don't have defined incident types
+
+fsa_data_3 <- fsa_data_3[-c(62:63),]
 view(fsa_data_3)
 
 ------------------------------------------------------------------------------------------
@@ -120,3 +125,4 @@ is.na(test_clean_map_data)
 
 ## Test for negative values within the whole dataframe
 test_clean_map_data <= 0
+

### 2024-09-27 paper/paper.qmd
Shapefile error - need to try again.

@@
 
-To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map the driver of EMS demand within each FSA.
-For the maps within this analysis, the analysis data explained in @sec-analysis needed to be modified.
-By grouping the data by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found.
+To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map EMS demand across all the locations Toronto Paramedic Services respond to.
+The map is based on the Paramedic Services Incident Data from 2022.
+By grouping the data shown in @sec-analysis by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found.
 The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls.
 A small snippet of the modified analysis data is shown below in @tbl-mapdata.
-The column that contains each FSA is named "CFSAUID" to ensure that each FSA can be mapped.
+The column, "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022.
 
@@
 
-To map this data, a shapefile with the boundaries for Toronto's FSAs was required.
-This data, "2021 Census - Boundary Files", was acquired from the Open Government Data portal by the Government of Canada [@2021census].
-Additional information about this dataset, including how it was prepared and used can be found in the Appendix (@sec-appendix).
+To map this data, a shapefile with the spatial information for all recorded FSAs is combined with the data shown in @tbl-mapdata.
+The shapefile, titled, "2021 Census - Boundary Files", is provided by the Open Government Data portal from the Government of Canada [@2021census].
+Additional information about this dataset, including how it is used can be found in the Appendix (@sec-appendix).
 
@@
 library(sf)
-library(repr)
+#library(repr)
 
@@
 
-toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas(TOR)")
-#view(toronto_forward_sortation_areas)
+forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas")
+view(forward_sortation_areas)
 
@@
 
-options(repr.plot.width = 20, repr.plot.height = 16)
+#options(repr.plot.width = 20, repr.plot.height = 16)
 

### 2024-09-27 paper/paper.qmd
finalized maps and continuing to edit writing.

@@
 
-Similarly, as vehicle accidents are a second potential driver of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident-related calls.
+Similarly, as vehicle accidents can be a second potential driver of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident-related calls.
 Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
@@
 By examining EMS demand through different incident types, the investigation above finds that medical emergencies are responsible for a significantly large portion of 911 calls.
-However, this finding applies to the city of Toronto in general, and emergency calls for medical emergencies may not be the driver of EMS demand within individual regions of Toronto.
-Thus, the following part of this analysis looks to identify the driver for each Forward Sortation Area (FSA) in Toronto.
-If different FSAs have different drivers of EMS demand, then strategies that are tailored to each region's circumstance may be more effective in managing EMS demand across the city.
+However, this finding is a generalization of all the forward sortation areas that Toronto Paramedic Services responded to over several years.
+Medical emergencies may not be the driver of EMS demand for all regions or communities.
+Thus, the following part of this analysis looks to identify the driver for every unique forward sortation area Toronto Paramedic Services responded to in 2022.
+If different forward sortation areas have different drivers of EMS demand, then strategies that are tailored to each region's circumstance may be more effective in managing overall EMS demand.
 
-It is important to note that there are only 96 FSAs that can be mapped for the city of Toronto [@2021census].
-For this reason, while the Paramedic Services Incident Data provides records of emergency calls for 184 unique FSAs, only 96 could be mapped in @fig-map below.
-So, although Toronto Paramedic Services have responded to calls throughout the Census Metropolitan Area (CMA) of Toronto and other surrounding regions, calls that occurred within the city of Toronto as outlined by the 2021 Census - Boundary Files data are considered.
+@fig-map illustrates the individual drivers of EMS demand for all forward sortation areas.
+From this illustration, clear differences in what propels EMS demand can be observed.
+For instance, in the larger forward sortation areas that surround the city of Toronto, incidents that require emergency transfers appear to be factors that raise EMS demand.
 
@@
 #| label: fig-map
-#| fig-cap: EMS Demand Across Toronto Forward Sortation Areas (Based on data from 2022)
+#| fig-cap: EMS Demand Across Forward Sortation Areas (Based on 2022 data)
 
@@
 library(sf)
-#library(repr)
 
@@
 
-forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas")
-view(forward_sortation_areas)
+forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
+#view(forward_sortation_areas)
+
+new_forward_sortation_areas <- st_transform(forward_sortation_areas, crs = 4326)
+#new_forward_sortation_areas
 
@@
 
-#ggplot(toronto_forward_sortation_areas) + 
+#ggplot(forward_sortation_areas) + 
   #geom_sf()
@@
 
-merged_map_data <- merge(toronto_forward_sortation_areas, map_data, by = "CFSAUID")
+merged_map_data <- merge(new_forward_sortation_areas, map_data, by = "CFSAUID")
 #view(merged_map_data)
@@
 # Map the merged data
-
-#options(repr.plot.width = 20, repr.plot.height = 16)
 
@@
        x = "Longitude", y="Latitude", 
-       caption = "Data Source: Open Canada 2021 Census & Toronto Paramedic Services") +
+       caption = "Data Source: 2021 Census & Toronto Paramedic Services") +
   theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
   scale_fill_discrete(name = "EMS Demand Driver") + 
-  scale_fill_manual(values = c("lightblue"))
+  scale_fill_manual(values = c("lightblue", "yellow", "lightpink")) +
+    coord_sf(xlim = c(-82, -76), ylim = c(42.5, 45.6), expand = FALSE)
+
 ```
 
-@fig-map illustrates the drivers of EMS demand for each FSA within the city of Toronto, and the key driver for all 96 FSAs in the city appears to be medical emergencies.
-This shows that for at least these 96 FSAs, strategies that focus on promoting health and well-being may be effective in reducing the pressure paramedic services face with high EMS demand.
-However, there are many more FSAs that paramedic services respond to and [TABLE] shows the remaining FSAs and their respective EMS demand drivers.
+```{r, results = 'hide'}
+#| message: false
+#| echo: false 
+#| warning: false
+#| label: fig-mapzoom
+#| fig-cap: A Closer Look at EMS Demand Across Forward Sortation Areas
 
-From [TABLE], it is evident that not all FSAs have the same driver for EMS demand.
-For instance, the FSA, "", has emergency transfers as its driver while "" has accidents as it driver.
-Thus, the takeaway from this analysis is that paramedic service strategies that are tailored to a specific region and its characteristics and needs, may be better.
-As each region or FSA will vary from others, this variability will make a "one size fits all approach" diffiuclt to employ.
-so, tailored strategies within paramedic services (allocation of resources, staff) as well as within each region (e.g.
-programs) are crucial.
+library(maps)
+library(sf)
+
+# Reading in the shapefile
+
+forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
+view(forward_sortation_areas)
+
+new_forward_sortation_areas <- st_transform(forward_sortation_areas, crs = 4326)
+#new_forward_sortation_areas
+
+# Reading in the paramedic services data
+
+map_data <- read_csv(here::here("data/analysis_data/analysis_data_FOR_MAP.csv"))
+#view(map_data)
+
+# Test to see if shapefile produces a map
+
+#ggplot(forward_sortation_areas) + 
+  #geom_sf()
+
+# Merge map_data and toronto_forward_sortation_areas by the column "CFSAUID"
+
+merged_map_data <- merge(new_forward_sortation_areas, map_data, by = "CFSAUID")
+#view(merged_map_data)
+
+# Map the merged data
+
+ggplot(merged_map_data) + 
+  geom_sf(aes(fill = EMS_Demand_Driver)) +
+  labs(title = "EMS Demand Drivers By Forward Sortation Area (FSA)",
+       x = "Longitude", y="Latitude", 
+       caption = "Data Source: 2021 Census & Toronto Paramedic Services") +
+  theme_light() + theme(plot.title = element_text(hjust = 0.3)) +
+  scale_fill_discrete(name = "EMS Demand Driver") + 
+  scale_fill_manual(values = c("lightblue", "yellow", "lightpink")) +
+    coord_sf(xlim = c(-80, -78.8), ylim = c(43.4, 44.3), expand = FALSE)
+
+```
 
@@
 
-The shapefile that contains the boundaries for Toronto's FSAs also contained the FSAs for every province and territory of Canada.
-As this analysis only considers the city of Toronto, ArcGIS Pro software [@arcgis] use employed to remove the spatial information of FSA locations outside Toronto.
-Specifically, the records of FSA regions that were not in Toronto were removed from the shapefile's attribute table (i.e.
-data table) using ArcGIS Pro.
-After modifying this data, it was merged with the data presented in @tbl-mapdata and mapped to show the "EMS Demand Driver" for each FSA.
+The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada.
+In this file, every forward sortation area is matched with a set of spatial information (i.e.
+coordinates).
+@tbl-shapefile illustrates this data.
+"CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have.
+The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively.
+"LANDAREA" is the total area in square kilometers of each forward sortation area, and "Geometry" holds the spatial information for every area.
+
+```{r, results = 'hide'}
+#| echo: false
+#| message: false
+#| label: tbl-shapefile
+#| tbl-cap: "Forward Sortation Areas Location Data"
+
+data <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
+
+data |>
+  slice(1:3) |>
+  kable(
+    col.names = c("CFSAUID", "DGUID", "PRUID", "PRNAME", 
+                  "LANDAREA", "Geometry"),
+    digits = 1,
+    booktabs = TRUE,
+    linesep = "",
+    align = c("l", "l", "l", "l", "l", "l"),
+    format.args = list(big.mark = ",")
+  )
+```
+
+To isolate the forward sortation areas given in the Paramedic Services Incident Data for 2022, ArcGIS Pro software [@arcgis] was used to remove any areas that were not recorded.
+Specifically, the forward sortation areas that were not present within the Paramedic Services Incident Data were deleted from the shapefile's attribute table (i.e.
+data table) and re-saved as a new shapefile.
+After making the necessary modifications, the spatial data for forward sortation areas was merged with the data present in @tbl-mapdata and mapped to show the "EMS Demand Driver" for every area.
 The final map is shown in @fig-map.

### 2024-09-27 paper/paper.qmd
Finished edits for entire paper - will review again as needed.

@@
 ---
-title: "Toronto Paramedic Services"
-subtitle: "A deeper look into what drives Emergency Medical Services (EMS) demand."
+title: "What Drives Emergency Medical Services (EMS) Demand?"
+subtitle: "Community-based EMS-demand-management strategies that focus on medical emergencies, emergency transfers, and vehicle accidents may help enhance paramedic services"
 author: 
@@
 date-format: long
-abstract: "In 2023, there were over 1,000 cases where an ambulance was unavailable to respond to an emergency call.
-This calls for urgent action as every moment is critical in emergency situations.
-By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand within Toronto to help inform more effective strategies for managing EMS demand.
-In short, medical emergencies and vehicle accidents appear to be the most prominent drivers of EMS demand, suggesting that strategies that seek to better manage demand in these areas may enhance the overall efficiency of paramedic services."
+abstract: "By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand to help inform demand-management strategies.
+Through data visualization, this analysis finds that medical emergencies, incidents requiring emergency transfers, and vehicle accidents appear to be the most prominent drivers of EMS demand.
+This suggests that strategies that target these drivers may enhance the overall efficiency of paramedic services.
+Additionally, the present analysis emphasizes that thoughtful consideration for the unique needs of those who require EMS is important to build practical and productive EMS-demand-management strategies."
 format: pdf
@@
 In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed.
-Thus, this paper seeks to investigate the factors that seem to be driving EMS demand within Toronto.
+Thus, this paper seeks to investigate the factors that seem to be driving EMS demand.
 Using Paramedic Services Incident Data offered by Toronto Paramedic Services, the aim is to identify which types of emergencies (e.g.
-medical emergencies or vehicle accidents) strongly drive EMS demand.
+medical emergencies or vehicle accidents) drive EMS demand and examine whether .
 Knowing this can help inform effective strategies that allocate resources to these drivers for better EMS demand management.
 
-With this, this paper first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time.
-Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to determine which of these five incident types are key drivers of the total EMS demand.
-As a result, this analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.
+With this, this analysis first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time.
+Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to identify the incident types that push EMS demand.
+Lastly, this analysis considers whether drivers of EMS demand differ between different regions.
+As a result, the present analysis finds that overall, medical emergencies, incidents requiring emergency transfers, and vehicle accidents seem to be the most prominent drivers of EMS demand.
+However, as different drivers of EMS demand are found for different regions, this analysis highlights that while strategies that target medical emergencies, incidents requiring emergency transfers, and vehicle accidents may enhance the overall efficiency of paramedic services, it is also crucial to consider the unique needs of various regions and their residents to shape stronger EMS-demand-management strategies.
 
-In what follows, the paper will define the data that was used within this analysis (@sec-data), provide a detailed account of the results (@sec-results), and put forth a discussion of the analysis along with its limitations and next steps (@sec-discussion).
+In what follows, a description of the data that was used within this analysis (@sec-data), detailed account of the results (@sec-results), and a brief discussion of the analysis along with its limitations and next steps (@sec-discussion) will be provided.
 
@@
 This pattern likely reflects the the onset of the covid-19 pandemic.
-Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS within Toronto.
+Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS.
 
@@
 By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
-With @fig-incidentemsdemandzoom, it becomes clear that a second driver of EMS demand may be vehicle accidents, though not as strong as medical emergencies.
+With @fig-incidentemsdemandzoom, it becomes clear that vehicle accidents and emergency transfers may be other potential drivers of EMS demand, though not as strong as medical emergencies.
 
@@
 #| label: fig-incidentemsdemandzoom
-#| fig-cap: EMS Demand for Each Year between 2017 and 2022
+#| fig-cap: Zooming in on Figure 2
 
@@
 
-As a prominent driver of EMS demand is medical-emergency-related calls, strategies that focus on efficiently managing calls for medical emergencies may help to better navigate EMS demand as a whole.
+As medical emergencies seem to be prominent drivers of EMS demand, strategies that focus on efficiently managing calls for medical emergencies may help to better navigate EMS demand as a whole.
 Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport].
@@
 
-Similarly, as vehicle accidents can be a second potential driver of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident-related calls.
+Similarly, as vehicle accidents and emergency transfers can be potential drivers of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident- and emergency-transfer-related calls.
 Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
-To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.
+To efficiently navigate EMS demand for vehicle accidents and emergency transfers, strategies may center around vehicle accident prevention or the allocation of more staff and resources to accident or emergency-prone regions for faster response times.
 
@@
 
+By focusing in on the city of Toronto with @fig-mapzoom, EMS demand seems to be significantly driven by medical emergencies and slightly driven by vehicle accidents and emergency transfers.
+
+These results suggest that what drives EMS demand is dependent on the characteristics of a particular region.
+Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident.
+Thus, there is a need to consider individual communities to build tailored solutions that effectively address their unique needs and better manage widespread EMS demand.
+
+\newpage
+
 # Discussion {#sec-discussion}
@@
 
-As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it. Thus, this analysis investigates...
+As high demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help to inform effective demand-management strategies.
+Through an investigation of EMS demand, the present analysis identifies three main drivers of EMS demand (medical emergencies, incidents requiring emergency transfers, and vehicle accidents) and also considers how these drivers might differ in different contexts (i.e.
+different regions or communities).
 
-To this, the given analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.
+To further this research, the relationship between the driver of EMS demand for every forward sortation area and each area's socio-economic or demographic characteristics should be examined.
+As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can help build tailored and effective strategies.
+It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular region.
 
-Thus, these findings offer a deeper understanding about the underlying factors that can drive EMS demand, and this understanding can further facilitate the exploration of various strategies that allow paramedic services to effectively operate even under high demand.
+In short, the findings of this analysis offer a deeper understanding about what drives EMS demand, and this understanding can further facilitate the exploration of various strategies that allow paramedic services to effectively operate even under high demand.
 
@@
 
-CONTENT
+Though this analysis carefully examines EMS demand, there are some limitations to consider.
+It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis.
+However, this may have limited the overall representation of each incident type as well as the conclusions that are made.
+Further, it is important to note that @fig-map only data from 2022.
+So, although @fig-map demonstrates that differences between regions and their drivers for EMS demand can exist, it is not an accurate reflection of what each forward sortation area because only one instance (the year 2022) is considered.
 
@@
 
-The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada.
-In this file, every forward sortation area is matched with a set of spatial information (i.e.
-coordinates).
-@tbl-shapefile illustrates this data.
-"CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have.
-The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively.
-"LANDAREA" is the total area in square kilometers of each forward sortation area, and "Geometry" holds the spatial information for every area.
-
-```{r, results = 'hide'}
+```{r}
 #| echo: false
 #| message: false
-#| label: tbl-shapefile
+#| warning: false
+#| label: tbl-shape
 #| tbl-cap: "Forward Sortation Areas Location Data"
 
-data <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
+library(sf)
+
+data <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas", quiet=TRUE)
 
@@
 ```
+The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada.
+In this file, every forward sortation area is matched with a set of spatial information (i.e.
+coordinates).
+@tbl-shape illustrates this data.
+"CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have.
+The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively.
+"LANDAREA" is the total area in square kilometers of each forward sortation area, and "Geometry" holds the spatial information for every area.
 

### 2024-09-27 scripts/03-data_cleaning_FOR_MAP.R
Didn't make any changes - accidentally pressed space bar

@@ -111,7 +111,7 @@ write_csv(fsa_data_3, file = "data/analysis_data/analysis_data_FOR_MAP.csv")
 # column
 
 test_clean_map_data <- read_csv("data/analysis_data/analysis_data_FOR_MAP.csv")
-#view(test_clean_map_data)
+#view(test_clean_map_data) 
 
 test_clean_map_data$`Emergency Transfers` == str_count(test_clean_map_data$Incident_Type, "Emergency Transfer")
 

### 2024-09-27 paper/paper.qmd
Edited the entire paper and rendered a new pdf version.

@@
 title: "What Drives Emergency Medical Services (EMS) Demand?"
-subtitle: "Community-based EMS-demand-management strategies that focus on medical emergencies, emergency transfers, and vehicle accidents may help enhance paramedic services"
+subtitle: "Community-based paramedic service strategies that focus on medical emergencies, emergency transfers, and vehicle accidents may help to effectively manage EMS demand"
 author: 
@@
 abstract: "By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand to help inform demand-management strategies.
-Through data visualization, this analysis finds that medical emergencies, incidents requiring emergency transfers, and vehicle accidents appear to be the most prominent drivers of EMS demand.
-This suggests that strategies that target these drivers may enhance the overall efficiency of paramedic services.
-Additionally, the present analysis emphasizes that thoughtful consideration for the unique needs of those who require EMS is important to build practical and productive EMS-demand-management strategies."
+Through data visualization, this analysis finds that medical emergencies, incidents requiring emergency transfers, and vehicle accidents are overall drivers of EMS demand.
+This suggests that centering strategies around these drivers may enhance the overall efficiency of paramedic services.
+Additionally, this analysis emphasizes that thoughtful consideration for the unique needs of those who require EMS is important to build practical EMS-demand-management strategies."
 format: pdf
@@
 However, research shows that this is not the case.
-A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 [@auditorreport].
-With drastically increasing wait times for those who required Emergency Medical Services (EMS), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
+A report on Toronto's paramedic services highlights that there were no ambulances available for over 1,000 calls in 2023 [@auditorreport].
+With drastically increasing wait times for those who required Emergency Medical Services (EMS), it is necessary to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.
 
 In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed.
-Thus, this paper seeks to investigate the factors that seem to be driving EMS demand.
-Using Paramedic Services Incident Data offered by Toronto Paramedic Services, the aim is to identify which types of emergencies (e.g.
-medical emergencies or vehicle accidents) drive EMS demand and examine whether .
-Knowing this can help inform effective strategies that allocate resources to these drivers for better EMS demand management.
+Thus, by using Paramedic Services Incident Data offered by Toronto Paramedic Services, this analysis seeks to identify which types of emergencies (e.g.
+medical emergencies or vehicle accidents) drive EMS demand.
+Knowing the drivers of EMS demand can help inform paramedic service strategies that allow for better EMS demand management.
 
 With this, this analysis first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time.
-Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to identify the incident types that push EMS demand.
-Lastly, this analysis considers whether drivers of EMS demand differ between different regions.
-As a result, the present analysis finds that overall, medical emergencies, incidents requiring emergency transfers, and vehicle accidents seem to be the most prominent drivers of EMS demand.
-However, as different drivers of EMS demand are found for different regions, this analysis highlights that while strategies that target medical emergencies, incidents requiring emergency transfers, and vehicle accidents may enhance the overall efficiency of paramedic services, it is also crucial to consider the unique needs of various regions and their residents to shape stronger EMS-demand-management strategies.
+Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to identify the drivers that push EMS demand.
+Lastly, this analysis considers whether drivers of EMS demand differ across different communities.
+As a result, the present analysis finds that overall, medical emergencies, incidents requiring emergency transfers, and vehicle accidents are the key drivers of EMS demand.
+However, not all communities are found to have the same drivers.
+So, while strategies that target medical emergencies, incidents requiring emergency transfers, and vehicle accidents may enhance the overall efficiency of paramedic services, it is important to consider the unique needs of various communities to shape stronger EMS-demand-management strategies.
 
-In what follows, a description of the data that was used within this analysis (@sec-data), detailed account of the results (@sec-results), and a brief discussion of the analysis along with its limitations and next steps (@sec-discussion) will be provided.
+In what follows, a description of the data that was used within this analysis (@sec-data), a detailed account of the results (@sec-results), and a brief discussion of the analysis (@sec-discussion) is provided.
 
@@
 The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls.
-A small snippet of the modified analysis data is shown below in @tbl-mapdata.
+This data is shown below in @tbl-mapdata.
 The column, "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022.
@@
 To map this data, a shapefile with the spatial information for all recorded FSAs is combined with the data shown in @tbl-mapdata.
-The shapefile, titled, "2021 Census - Boundary Files", is provided by the Open Government Data portal from the Government of Canada [@2021census].
+The shapefile from the 2021 Census is provided by the Open Government Data portal from the Government of Canada [@2021census].
 Additional information about this dataset, including how it is used can be found in the Appendix (@sec-appendix).
@@
 
-## EMS Demand from 2017 to 2022 Based on Incident Type
+## EMS Demand from 2017 to 2022 Based on Incident Type {#sec-incidenttypes}
 
-To further unpack what is driving this consistent demand, the calls for individual incident type are also considered.
+To further unpack what is driving this consistent demand, the number of emergency calls for individual incident type are also considered.
 @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls.
@@
 
-Similarly, as vehicle accidents and emergency transfers can be potential drivers of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident- and emergency-transfer-related calls.
+Similarly, as vehicle accidents and emergency transfers can be potential drivers of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle accident and emergency transfer-related calls.
 Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected.
@@
 However, this finding is a generalization of all the forward sortation areas that Toronto Paramedic Services responded to over several years.
-Medical emergencies may not be the driver of EMS demand for all regions or communities.
-Thus, the following part of this analysis looks to identify the driver for every unique forward sortation area Toronto Paramedic Services responded to in 2022.
-If different forward sortation areas have different drivers of EMS demand, then strategies that are tailored to each region's circumstance may be more effective in managing overall EMS demand.
+Medical emergencies may not be the main driver of EMS demand for all communities.
+Thus, the following part of this analysis looks to identify the driver for every unique forward sortation area Toronto Paramedic Services have responded to in 2022.
+If different forward sortation areas have different drivers of EMS demand, then strategies that are tailored to each community's circumstance may be more effective in managing overall EMS demand.
 
@@
 #| label: fig-map
-#| fig-cap: EMS Demand Across Forward Sortation Areas (Based on 2022 data)
+#| fig-cap: EMS Demand Across Forward Sortation Areas (Based on 2022 Data)
 
@@
 forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
-view(forward_sortation_areas)
+#view(forward_sortation_areas)
 
@@
 
-By focusing in on the city of Toronto with @fig-mapzoom, EMS demand seems to be significantly driven by medical emergencies and slightly driven by vehicle accidents and emergency transfers.
+By focusing in on the city of Toronto with @fig-mapzoom, EMS demand in many communities seems to be significantly driven by medical emergencies, and there are only a few communities whose EMS demand is driven by vehicle accidents or emergency transfers.
 
-These results suggest that what drives EMS demand is dependent on the characteristics of a particular region.
+These results support the findings from @sec-incidenttypes.
+Medical emergencies, emergency transfers, and vehicle accidents are shown to be relevant drivers of EMS demand.
+The findings further imply that what drives EMS demand is dependent on the characteristics of a particular region.
 Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident.
-Thus, there is a need to consider individual communities to build tailored solutions that effectively address their unique needs and better manage widespread EMS demand.
+Given this, general strategies that target the three main drivers of EMS demand may not be effective for all communities.
+Thus, there is a need to consider the characteristics of different communities to build community-based strategies that effectively address the main drivers of EMS demand.
 
@@
 As high demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help to inform effective demand-management strategies.
-Through an investigation of EMS demand, the present analysis identifies three main drivers of EMS demand (medical emergencies, incidents requiring emergency transfers, and vehicle accidents) and also considers how these drivers might differ in different contexts (i.e.
-different regions or communities).
+Through an investigation of EMS demand, the present analysis identifies three main drivers of EMS demand (medical emergencies, incidents requiring emergency transfers, and vehicle accidents) and also considers how these drivers might differ across various communities.
 
-To further this research, the relationship between the driver of EMS demand for every forward sortation area and each area's socio-economic or demographic characteristics should be examined.
-As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can help build tailored and effective strategies.
-It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular region.
+To further this research, the relationship between the driver of EMS demand and socio-economic or demographic characteristics for each forward sortation area should be examined.
+As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can allow for the formation of effective community-based strategies.
+It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular community.
 
@@
 It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis.
-However, this may have limited the overall representation of each incident type as well as the conclusions that are made.
-Further, it is important to note that @fig-map only data from 2022.
-So, although @fig-map demonstrates that differences between regions and their drivers for EMS demand can exist, it is not an accurate reflection of what each forward sortation area because only one instance (the year 2022) is considered.
+This may have limited the overall representation of the number of emergency calls for each incident type.
+Further, it is important to note that @fig-map only considers data from 2022.
+So, although @fig-map demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e.
+from 2017 to 2022).
 
@@
     linesep = "",
-    align = c("l", "l", "l", "l", "l", "l"),
+    align = c("l", "l", "l", "l", "l", "c"),
     format.args = list(big.mark = ",")
@@
 The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada.
-In this file, every forward sortation area is matched with a set of spatial information (i.e.
+In this file, every forward sortation area is matched with spatial information (i.e.
 coordinates).

### 2024-09-27 paper/paper.qmd
Reviewed and edited entire paper again.

@@
 Thus, by using Paramedic Services Incident Data offered by Toronto Paramedic Services, this analysis seeks to identify which types of emergencies (e.g.
-medical emergencies or vehicle accidents) drive EMS demand.
+medical emergencies, vehicle accidents, etc.) drive EMS demand.
 Knowing the drivers of EMS demand can help inform paramedic service strategies that allow for better EMS demand management.
@@
 Paramedic Services Incident Data (2017-2022) is provided by Toronto Paramedic Services [@torontoparamedicservices].
-The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all updated by hand to the city of Toronto's Open Data portal [@opentorontodata].
+The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all periodically updated by hand to the city of Toronto's Open Data portal [@opentorontodata].
 
@@
 
-The Paramedic Services Incident Data is refreshed annually and contains information on every single incident paramedics have responded to since January 1st to December 31st in 2017 to 2022.
+The Paramedic Services Incident Data is refreshed annually and contains information on every single incident paramedics have responded to since January 1, 2017 to December 31, 2022.
 For every day of each year, there are multiple records of incidents that have occurred.
@@
 
-To further unpack what is driving this consistent demand, the number of emergency calls for individual incident type are also considered.
+To further unpack what is driving this EMS demand, the number of emergency calls for individual incident type are also considered.
 @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls.
@@
 
-By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
+By zooming into the lower portion of @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided.
 With @fig-incidentemsdemandzoom, it becomes clear that vehicle accidents and emergency transfers may be other potential drivers of EMS demand, though not as strong as medical emergencies.
@@
 
-## Mapping EMS Demand
+## Mapping EMS Demand {#sec-maplocation}
 
@@
 Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident.
-Given this, general strategies that target the three main drivers of EMS demand may not be effective for all communities.
+Given this, general, blanketing strategies that target the three main drivers of EMS demand may not be effective for all communities.
 Thus, there is a need to consider the characteristics of different communities to build community-based strategies that effectively address the main drivers of EMS demand.
@@
 As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can allow for the formation of effective community-based strategies.
-It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular community.
+It may also be helpful to consider ambulance station locations in addition to @fig-map in @sec-maplocation to see whether ambulance stations and resources are optimally distributed and if additional resources are needed based on the EMS demand driver for a particular community.
 
@@
 Further, it is important to note that @fig-map only considers data from 2022.
-So, although @fig-map demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e.
+So, although @fig-map in @sec-maplocation demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e.
 from 2017 to 2022).
@@
 After making the necessary modifications, the spatial data for forward sortation areas was merged with the data present in @tbl-mapdata and mapped to show the "EMS Demand Driver" for every area.
-The final map is shown in @fig-map.
+The final map is shown in @fig-map in @sec-maplocation.
 

### 2024-09-27 paper/paper.qmd
Final revisions made to the entire paper!

@@
 Paramedic Services Incident Data (2017-2022) is provided by Toronto Paramedic Services [@torontoparamedicservices].
-The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all periodically updated by hand to the city of Toronto's Open Data portal [@opentorontodata].
+The records of when a paramedic was dispatched having received a call, the type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all periodically updated to the city of Toronto's Open Data portal [@opentorontodata].
 
@@
 The total number of emergency calls appears to increase in 2020 after a notable decrease in 2019.
-This pattern likely reflects the the onset of the covid-19 pandemic.
+This pattern likely reflects the onset of the covid-19 pandemic.
 Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS.
@@
 
-To further unpack what is driving this EMS demand, the number of emergency calls for individual incident type are also considered.
+To further unpack what is driving this EMS demand, the number of emergency calls for each incident type are also considered.
 @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls.
@@
 This may have limited the overall representation of the number of emergency calls for each incident type.
-Further, it is important to note that @fig-map only considers data from 2022.
-So, although @fig-map in @sec-maplocation demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e.
+Further, it is important to note that @fig-map in @sec-maplocation only considers data from 2022.
+So, although @fig-map demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e.
 from 2017 to 2022).
@@
 @tbl-shape illustrates this data.
-"CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have.
+"CFSAUID" contains every FSA in Canada, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have.
 The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively.

### 2024-09-27 scripts/00-simulate_data.R
Edited preamble for simulation script.

@@ -4,7 +4,7 @@
 # Date: 23 September 2024
 # Contact: jlee.lee@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Review the plan and sketches for the analysis in the "other" folder
+# Pre-requisites: Review the plan and sketches for the analysis in the "plan" folder
 # Any other information needed? N/A
 
 
